---
title: "Completeness Map of Demographics Data"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r, include=FALSE}
library(tidyverse)
library(naniar)
library(here)

data_path <- here("data")
```

```{r, warning=FALSE, message=FALSE, include=FALSE}
subjfilters <- fs::dir_ls(here("data", "sub_lists"))
ids <- subjfilters %>% 
  map_dfr(read_csv, .id = "source", col_names = c("participant_id")) %>%
  mutate(dataset = str_extract(source, "[a-z]+(?=_subj_list\\.csv)"))
```

In this notebook we will investigate the completeness of demographic data in the RBC datasets. We'll be basing this off of the notebook at `r here("notebooks", "RBC_PFactor_processing.Rmd")` from the Brazil team.

# HBN Processing
<!-- HBN -->

```{r, cache=TRUE, include=FALSE}
hbn_raw <- read_csv(here(data_path, "HBNdata_2020_12_03.csv")) %>%
  slice(-1) %>%
  rename("participant_id" = Identifiers) %>%
  mutate(participant_id = str_remove(participant_id, ",assessment")) %>%
  mutate(participant_id = str_c("sub-", participant_id))
```


```{r, warning=FALSE}
hbn_ <- hbn_raw %>%
  rename("age" = `Basic_Demos,Age`)
hbn_ <- hbn_ %>% 
  mutate(age = as.numeric(age)) %>%
  mutate(age_group = 
           case_when(
             age>=5 & age<11 ~ "5_to_10", 
             age>=11& age<15 ~ "11_to_14",
             age>=15 ~ "15_or_higher"
             )
         ) %>%
  mutate(age_group = ordered(age_group, 
                             levels = c("5_to_10", "11_to_14", "15_or_higher")))
```

<!-- Gender -->

```{r}
hbn_ <- hbn_ %>%
  rename(sex = `Basic_Demos,Sex`) %>%
  mutate(sex = case_when(sex=="0" ~ "Male", 
                            sex=="1" ~ "Female"
                            ) %>% as.factor())
```

<!-- Study Site -->

```{r}
hbn_ <- hbn_ %>%
  rename(study_site = `Basic_Demos,Study_Site`) %>%
  mutate(study_site = case_when(study_site =="1" ~ "HBN-1",
                                study_site =="2" ~ "HBN-2",
                                study_site =="3" ~ "HBN-3",
                                study_site =="4" ~ "HBN-4",
                                study_site =="5" ~ "HBN-5"),
         study = "HBN")
```

<!-- Race -->

```{r, warning=FALSE}
race_levels <- seq(0:11)
race_labels <- c('White', 'Black', 'Mixed', 'Asian', 'Asian', 'Native', 'Native', 'Native', 'Mixed', 'Other', 'Unknown', 'Other')
race_labels_old <- c("White/Caucasian", "Black/African American", "Hispanic", "Asian", "Indian", "Native American Indian", "American Indian/Alaskan Native", "Native Hawaiian/Other Pacific Islander", "Two or More Races", "Unknown", "Other", "Other")

hbn_ <- hbn_ %>%
  #rename(race=`PreInt_Demos_Fam,Child_Race`) %>%
  mutate(race = factor(as.numeric(`PreInt_Demos_Fam,Child_Race`), levels=race_levels, labels=race_labels))
#0= White/Caucasian
#1= Black/African American
#2= Hispanic
#3= Asian
#4= Indian
#5= Native American Indian
#6= American Indian/Alaskan Native
#7= Native Hawaiian/Other Pacific Islander
#8= Two or more races
#9= Other race
#10= Unknown

# "0='White';1='Black'; 2='Mixed';3='Asian';4='Asian'; 5='Native'; 6='Native'; 7='Native'; 8='Mixed'; 9='Other'; 10='Unknown'; 11='Other'"
```

<!-- BMI -->


```{r, warning=FALSE}
hbn_ <- hbn_ %>%
  mutate(bmi=as.numeric(`Physical,BMI`))
```


<!-- Education -->
```{r, warning=FALSE}
hbn_ <- hbn_ %>%
  mutate(education_years=as.numeric(`PreInt_EduHx,yrs_school`))
```


```{r}

```


<!--END HBN-->

# NKI Processing

```{r, cache=TRUE, include=FALSE}
 # NKI
nki_raw1 <- read_csv(here(data_path, "RocklandSample_04JAN2021.csv")) %>%
  rename("participant_id" = Identifiers) %>%
  separate(participant_id, c("participant_id", "wave"))

nki_raw2 <- read_csv(here(data_path, "RocklandSample_01SET2021_cog.csv")) %>%
  rename("participant_id" = Identifiers) %>%
  separate(participant_id, c("participant_id", "wave"))

```

```{r, warning=FALSE}
nki_wave_levels <- c(1, 1, 1, 2, 3, 3)
nki_wave_labels <- c('BAS1', 'BAS2', 'BAS3', 'FLU1', 'FLU2', 'FLU3')

nki_ <- left_join(nki_raw1, nki_raw2, by=c("participant_id", "wave")) %>%
  mutate(participant_id = str_c("sub-", participant_id))
nki_ <- nki_ %>%
  mutate(
    wave = case_when(
      wave == "BAS1"    ~ 1,
      wave == "BAS2"    ~ 1,
      wave == "BAS3"    ~ 1,
      wave == "FLU1"    ~ 2,
      wave == "FLU2"    ~ 3,
      wave == "FLU3"    ~ 3
      ) %>% factor(levels = nki_wave_levels, labels =nki_wave_labels)
    )

```

```{r, warning=FALSE}
nki_race_old_levels <- c(1:6)
nki_race_old_labels <- c("American Indian or Native Alaskan", "Asian", "Black  or African American", "Native Hawiian or Other Pacific Islander", "White", "Other")

nki_ <- nki_ %>%
  
  # study data
  mutate(study = "NKI",
         study_site = "NKI-1",
         selection = "Community"
  ) %>%
  
  # age as above
  rename(age = nkicbcl_128) %>%
  mutate(age = as.numeric(age),
         age_group = 
           case_when(
             age>=5 & age<11 ~ "5_to_10", 
             age>=11& age<15 ~ "11_to_14",
             age>=15 ~ "15_or_higher"
             )
         ) %>%
  mutate(age_group = ordered(age_group, 
                             levels = c("5_to_10", "11_to_14", "15_or_higher"))) %>%
  
  # gender as above
  mutate(sex = 
           case_when(
             nkicbcl_127=="1" ~ "Male",
             nkicbcl_127=="2"~ "Female")
         ) %>%
  
  # race as above
  mutate(race_old = factor(dem_004, labels = nki_race_old_labels, levels = nki_race_old_levels),
         race = 
           case_when(
             dem_004 == 1    ~ "Native",
             dem_004 == 2    ~ "Asian",
             dem_004 == 3    ~ "Black",
             dem_004 == 4    ~ "Native",
             dem_004 == 5    ~ "White",
             dem_004 == 6    ~ "Other"
           ) %>% factor()
         ) %>%
  
#NKI_data <- NKI_data %>% mutate(Child_race=car::recode(Child_race,"5='White';3='Black'; 2='Asian'; 1='Native'; 4='Native'; 6='Other'"))
#1= American Indian or Native Alaskan
#2= Asian
#3= Black  or African American
#4= Native Hawiian or Other Pacific Islander
#5= White
#6= Other Race
  
  # Education
  mutate(education_years = as.numeric(sesc_10))
```

# PNC Processing
<!-- PNC -->

```{r, message=FALSE, warning=FALSE}
pnc_raw1 <- read_csv(here("data", "pnc_data.csv")) %>% 
  rename(participant_id=bblid)
pnc_raw2 <- read_csv(here("data", "bblid_renamed.csv")) %>%
      rename(participant_id=reid)
pnc_data <- pnc_raw1  %>%
  left_join(
    pnc_raw2
    ) %>%
  mutate(participant_id = str_c("sub-", participant_id))
```

```{r}
pnc_ <- pnc_data %>%
  rename(wave=timepoint) %>%
  mutate(study = "PNC",
         selection = "Community",
         study_site = "PNC-1") %>%
  mutate(sex = 
           case_when(sex=="1" ~ "Male", sex=="2"~ "Female") %>% as.factor()
         ) %>%
  mutate(
    age = as.numeric(ageAtClinicalAssess1)/12,
    age_group = 
      case_when(
        age>=5 & age<11 ~ "5_to_10", 
        age>=11& age<15 ~ "11_to_14",
        age>=15 ~ "15_or_higher"
        )
    ) %>%
  mutate(age_group = ordered(age_group, 
                             levels = c("5_to_10", "11_to_14", "15_or_higher"))) %>%
  mutate(education_years = edu1) %>%
  
  #PNC_data <- PNC_data %>% mutate(Child_race=car::recode(Child_race,"1='White'; 2='Black'; 3='Native'; 5='Mixed'; 4='Asian'; 6='Native'"))
  mutate(race = factor(race,
                       levels = seq(1:6),
                       labels = c("White", "Black", "Native", "Mixed", "Asian", "Native")
                       )
         ) %>%
  mutate(study = "PNC") %>%
  mutate(father_edu = fedu1, mother_edu = medu1, handedness = handednessv2)

```


<!-- END PNC-->

# HRC Processing

```{r, cache=TRUE, warning=FALSE}
# HRC
BHRC_data <- readRDS(here("data", "BHRC_2020_11_05.rds"))

# correction to include site, gender and skincolor in all waves and filter for only those with responses at follow-ups
mydata_w0 <- BHRC_data %>% dplyr::filter(redcap_event_name=="wave0_arm_1", dcany!="NA")
mydata_w1 <- BHRC_data %>% dplyr::filter(redcap_event_name=="wave1_arm_1", dcany!="NA")
mydata_w2 <- BHRC_data %>% dplyr::filter(redcap_event_name=="wave2_arm_1", dcany!="NA") %>% dplyr::select(-state, -gender, -skincolor)
mydata_complement <- BHRC_data %>% dplyr::filter(redcap_event_name=="wave0_arm_1") %>% dplyr::select(ident, state, gender, skincolor)
mydata_w2 <- left_join(mydata_w2,mydata_complement, by="ident") 
BHRC_data <- rbind(mydata_w0, mydata_w1, mydata_w2)

BHRC_data<-dplyr::rename(BHRC_data,EID=ident); BHRC_data$EID<-as.character(BHRC_data$EID)
BHRC_data$Study<-"BHRC"
BHRC_data<-BHRC_data %>% mutate(selection = case_when(selection==1 ~ "Community", 
                                       selection==2 ~ "High_risk"))
BHRC_data<-BHRC_data %>% mutate(wave = case_when(redcap_event_name=="wave0_arm_1" ~ 1, 
                                       redcap_event_name=="wave1_arm_1" ~ 2,
                                       redcap_event_name=="wave2_arm_1" ~ 3))
#BHRC_data<-BHRC_data %>% mutate(Site = case_when(state=="Rio Grande do Sul" ~ "BHRC-1", state=="São Paulo"~ "BHRC-2"))
BHRC_data<-BHRC_data %>% mutate(Site = case_when(state=="1" ~ "BHRC-1", state=="2"~ "BHRC-2"))
#BHRC_data<-BHRC_data %>% mutate(Gender = case_when(gender=="Female" ~ 1, gender=="Male"~ 2))
BHRC_data<-BHRC_data %>% mutate(sex = case_when(gender=="1" ~ "Male", gender=="2"~ "Female"))
BHRC_data$gender <- as.factor(BHRC_data$sex)
BHRC_data<-dplyr::rename(BHRC_data,Age=nage)
BHRC_data<-BHRC_data %>% mutate(Age_group_3cat = case_when(Age>=5 & Age<11 ~ "5_to_10", 
                                                      Age>=11& Age<15 ~ "11_to_14",
                                                      Age>=15 ~ "15_or_higher"))
BHRC_data$Age_group_3cat <- ordered(BHRC_data$Age_group_3cat, levels = c("5_to_10", "11_to_14", "15_or_higher")) 

BHRC_data$Child_race <- BHRC_data$skincolor
#BHRC_data <- BHRC_data %>% mutate(Child_race=car::recode(Child_race,"1='White';2='Black'; 3='Mixed';5='Asian'; 4='Native'"))

#education
BHRC_data_w0 <- BHRC_data %>% dplyr::filter(wave==1)
BHRC_data_w1 <- BHRC_data %>% dplyr::filter(wave==2)
BHRC_data_w2 <- BHRC_data %>% dplyr::filter(wave==3)

BHRC_data_w0 <- BHRC_data_w0 %>% mutate(School_repet= 
                case_when(fat24==1 ~ 1,
                          fat24==0 ~ 0,
                          fat24==99 ~ NA_real_))
BHRC_data_w0 <- BHRC_data_w0 %>% mutate(School_exp= 
                case_when(fat25==1|fat26==1 ~ 1,
                          fat25==0 & fat26==0 ~ 0,
                          fat25==99 ~ NA_real_))
BHRC_data_w0 <- BHRC_data_w0 %>% mutate(School_susp= 
                case_when(fat27==1 ~ 1,
                          fat27==0 ~ 0,
                          fat27==99 ~ NA_real_))

BHRC_data_w1 <- BHRC_data_w1 %>% mutate(School_repet= 
                case_when(le4_unite==1 ~ 1,
                          le4_unite==0 ~ 0,
                          le4_unite==99 ~ NA_real_))
BHRC_data_w1 <- BHRC_data_w1 %>% mutate(School_exp= 
                case_when(le5_unite==1|le7_unite==1|instlevel==13 ~ 1,
                          le5_unite==0 & le7_unite==0 & instlevel!=13 ~ 0,
                          le5_unite==99 & le7_unite==99 ~ NA_real_))
BHRC_data_w1 <- BHRC_data_w1 %>% mutate(School_susp= 
                case_when(le6_unite==1 ~ 1,
                          le6_unite==0 ~ 0,
                          le6_unite==99 ~ NA_real_))

BHRC_data_w2 <- BHRC_data_w2 %>% mutate(School_repet= 
                case_when(le4_unite==1 ~ 1,
                          le4_unite==0 ~ 0,
                          le4_unite==99 ~ NA_real_))
BHRC_data_w2 <- BHRC_data_w2 %>% mutate(School_exp= 
                case_when(le5_unite==1|le7_unite==1 ~ 1,
                          le5_unite==0 & le7_unite==0 ~ 0,
                          le5_unite==99 & le7_unite==99 ~ NA_real_))
BHRC_data_w2 <- BHRC_data_w2 %>% mutate(School_susp= 
                case_when(le6_unite==1 ~ 1,
                          le6_unite==0 ~ 0,
                          le6_unite==99 ~ NA_real_))

BHRC_data <- rbind(BHRC_data_w0, BHRC_data_w1, BHRC_data_w2)
rm(BHRC_data_w0, BHRC_data_w1, BHRC_data_w2)

BHRC_data<-dplyr::rename(BHRC_data,School_years=instlevel)
# summary.factor(BHRC_data$School_years)
# get_labels(BHRC_data$School_years)
# BHRC_data <- BHRC_data %>% mutate(School_years=car::recode(School_years,"13=NA"))
```


```{r}
bhrc_race_levels <- seq(1:5)
bhrc_race_labels <- c("White", "Black", "Mixed", "Native", "Asian")
bhrc_ <- BHRC_data %>%
  mutate(race = factor(Child_race, levels=bhrc_race_levels, labels=bhrc_race_labels))

bhrc_ <- bhrc_ %>%
  rename(participant_id=subjectid, age=Age, age_group=Age_group_3cat, race=race, education_years=School_years, study=Study, study_site=Site, wave=wave, sex=sex) %>%
  mutate(participant_id = str_c("sub-", participant_id))
```


# HCP-D

```{r, cache=TRUE, message=FALSE, warning=FALSE}
hcpd_raw <- read_csv(here("data", "hcpd_demographics.csv"))
```

```{r}
hcpd_raw
```






# HBN Summary

```{r, results='asis'}
hbn_final <- hbn_ %>%
  select(participant_id, age, sex, study, study_site, race, bmi, education_years) %>%
  filter(participant_id %in% ids$participant_id)

cat("Total subjects: ", length(unique(hbn_final$participant_id)))
cat("\n")
cat("Current Variables:\n")
cat(paste0("- ", colnames(hbn_final)), sep="\n")
```

```{r}
hbn_final %>%
  gg_miss_var(show_pct=TRUE) +
  labs(title="Missingness of Demographic Variables in HBN")
```

```{r, eval=FALSE}
hbn_ %>%
  select(age) %>%
  summary() %>%
  knitr::kable()
```

```{r, warning=FALSE, include=FALSE, eval=FALSE}
hbn_ %>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth = 2)+
  theme_minimal()
```

```{r, eval=FALSE, include=FALSE}
hbn_ %>%
  ggplot(aes(gender)) +
  geom_bar() +
  theme_minimal()
```

```{r, include=FALSE, eval=FALSE}
hbn_ %>%
  ggplot(aes(x=study_site)) +
  geom_bar() +
  theme_minimal()
```

```{r, warning=FALSE, include=FALSE, eval=FALSE}
hbn_ %>%
  ggplot(aes(x=race))+
  geom_bar() +
  theme_minimal() +
  ggtitle("Raw Count of Participant Race")
```

```{r, warning=FALSE, eval=FALSE, include=FALSE}
hbn_ %>%
  filter(!is.na(race)) %>%
  ggplot(aes(x=race))+
  geom_bar() +
  theme_minimal() +
  ggtitle("Raw Count of Participant Race (NA removed)")
```

```{r, warning=FALSE, include=FALSE, eval=FALSE}
hbn_ %>%
  ggplot(aes(x=bmi))+
  geom_histogram(binwidth = 2) +
  theme_minimal()
```

```{r, warning=FALSE, include=FALSE, eval=FALSE}
hbn_ %>%
  ggplot(aes(x=education))+
  geom_bar() +
  xlab("Education in years") +
  theme_minimal()
```

```{r, warning=FALSE, include=FALSE, eval=FALSE}
hbn_ %>%
  mutate(race_old = factor(as.numeric(`PreInt_Demos_Fam,Child_Race`), levels=race_levels, labels=race_labels_old)) %>%
  select(race, race_old) %>%
  ggplot(aes(x=race)) +
  geom_bar(aes(fill=race_old), position="fill") +
  theme_minimal() +
  ggtitle("Proportion of Old Race Labels Converted to New Labels") +
  ylab("proportion")
```


```{r, warning=FALSE}
hbn_final %>%
  gg_miss_var(show_pct=TRUE)+
  labs(title="Missingness of Demographic Variables in HBN")
```

# NKI Summary

```{r}
nki_final <- nki_ %>%
  select(participant_id, age, age_group, sex, race_old, race, education_years, study, study_site, selection) %>%
  filter(participant_id %in% ids$participant_id)# %>%
  #mutate(gender=factor(gender))
```

```{r}
cat("Total subjects: ", length(unique(nki_final$participant_id)))
cat("\n")
cat("Current Variables:\n")
cat(paste0("- ", colnames(nki_final)), sep="\n")
```

```{r, warning=FALSE}
nki_final %>%
  select(-race_old) %>%
  gg_miss_var(show_pct=TRUE)+
  labs(title="Missingness of Demographic Variables in NKI")
```

# PNC Summary


```{r}
pnc_final <- pnc_ %>%
  select(participant_id, age, age_group, sex, race, education_years, study, study_site, selection, wave) %>%
  filter(participant_id %in% ids$participant_id)
```

```{r}
cat("Total subjects: ", length(unique(pnc_final$participant_id)))
cat("\n")
cat("Current Variables:\n")
cat(paste0("- ", colnames(pnc_final)), sep="\n")
```


```{r}
pnc_final %>%
  gg_miss_var(show_pct=TRUE) +
  labs(title="Missingness of Demographic Variables in PNC")
```

# HRC Summary

```{r}
bhrc_final <- bhrc_ %>%
  select(participant_id, age, age_group, sex, race, education_years, study, study_site, selection, wave) %>%
  filter(participant_id %in% ids$participant_id)# %>%
  #mutate(gender = factor(Gender)) %>%
  #select(-Gender)
```


```{r}
cat("Total subjects: ", length(unique(bhrc_final$participant_id)))
cat("\n")
cat("Current Variables:\n")
cat(paste0("- ", colnames(bhrc_final)), sep="\n")
```

```{r}
bhrc_final %>%
  gg_miss_var(show_pct=TRUE) +
  labs(title="Missingness of Demographic Variables in HRC")
```

# Harmonization

```{r}
harm <- bind_rows(hbn_final, nki_final, pnc_final, bhrc_final)

harm <- harm %>%
  mutate(race = case_when(
    race == "White"       ~ "White",
    race == "Black"       ~ "Black",
    race == "Mixed"       ~ "Mixed",
    race == "Native"       ~ "Other",
    race == "Asian"       ~ "Asian",
    race == "Other"       ~ "Other"
  ) %>% as.factor())

harm %>%
  select(-race_old, -bmi, -age_group) %>%
  gg_miss_var(show_pct = TRUE)
```

## Variables Viz.

```{r}
harm %>%
  ggplot(aes(x=age))+
  geom_density(aes(fill=study), alpha=0.25)+
  theme_minimal() +
  ggtitle("Distribution of Age")
```


```{r}
harm %>%
  ggplot(aes(x=sex))+
  geom_bar()+
  theme_minimal() +
  ggtitle("Participat Sex")
```

# Discussion

- Done:
  - Collapsed race values

```{r, include=FALSE, warn=FALSE}
hbn_ %>%
  filter(!is.na(race)) %>%
  ggplot(aes(x=race))+
  geom_bar() +
  theme_minimal() +
  ggtitle("Raw Count of Participant Race in HBN (NA removed)")
```

```{r, include=FALSE}
hbn_ %>%
  mutate(race_old = factor(as.numeric(`PreInt_Demos_Fam,Child_Race`), levels=race_levels, labels=race_labels_old)) %>%
  select(race, race_old) %>%
  ggplot(aes(x=race)) +
  geom_bar(aes(fill=race_old), position="fill") +
  theme_minimal() +
  ggtitle("Proportion of Old Race Labels Converted to New Labels in HBN") +
  ylab("proportion")
```

```{r, include=TRUE, warning=FALSE}
harm %>%
  #filter(!is.na(race)) %>%
  ggplot(aes(x=race))+
  geom_bar() +
  theme_minimal() +
  ggtitle("Raw Count of Participant Race in Harmonized")
```

  - Sex vs. Gender

  - Collapsing of NKI *sessions* into *waves*

```
wave = case_when(
      wave == "BAS1"    ~ 1,
      wave == "BAS2"    ~ 1,
      wave == "BAS3"    ~ 1,
      wave == "FLU1"    ~ 2,
      wave == "FLU2"    ~ 3,
      wave == "FLU3"    ~ 3
      )
```

```
> nki_$wave %>% table()
.
BAS1 BAS2 BAS3 FLU1 FLU2 FLU3  NFB 
1484   92    1  492  351   23    4 
```

  - One participant with BAS3 and 4 with NFB -- how to deal?

  - What should we use for age NKI? Old value was `nkicbcl_128 -> age`
```{r}
nki_ %>%
  select(age, age_04) %>%
  mutate(across(everything(), as.numeric)) %>%
  summary()
```

- To-do:
  - HCP-D (data available)
  - CCNP (data available)
  - PACCT? (On-hold)
  - handedness
  - parent education
  - bmi
  - age at X





